PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2023-03-10 reviewer_response.tex

\reviewer{R2.7}{Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?}
% \Marco{the effect of merging more blocks on L50 is not trivial. It depends on whether the merging occurs between long or short blocks, and how much fragmentation this causes. HP is a special case in that its diverge is on the threshold of what minimap2 can handle. Using mmseqs2 results in a pangenome that is half the size compared to standard parameters (fig. 4G). This is an impressive amount of additional compression. At the same time the fraction of core pangenome more than doubles (4H). The total number of blocks decreases to roughly half (4D) while N50 increases (4F). For me this points to the fact that in the case of HP merging will split long blocks and create more fragmentation. For the reviewer I could produce the block frequency and lenght distribution for the two cases and comparing them would give an answer. I could do the same for EC or KP for comparison.}\\
% \Liam{makes sense.}
\response{This is an interesting observation. L50 measures the minimum number of pancontigs needed to cover at least 50\% of the pangenome graph sequence. 
This number is closely related to the total number of pancontigs. 
In general the use of null energy parameters ($\beta=\alpha=0$) results in more fragmented graphs, with a higher number of pancontigs (cf. fig.~{\SIfigBenchmark}D in SI) so L50 increases. 
However, the reviewer is correct that \textit{Helicobacter} is an exception and shows the opposite trend: the use of null energy parameters results in less fragmentation of the graph (both in the number of pancontigs and for the value of L50).\\
Having investigated, we believe this might be explained by that fact that the divergence of some parts of the \textit{Helicobacter} core genome in this dataset is close to $1/\beta$ for $\beta=10$ used in the figure.
Hence, parts of the core genome where the divergence reaches $>1/\beta$ might be split into different pancontigs. 
Removing the threshold imposed by $\beta$ by setting $\beta=0$ results in a longer consecutive merger and therefore a less fragmented graph with fewer overall pancontigs.\\
For a visual verification of this interpretation we display in fig.~\ref{fig:kp-vs-hp} how the distributions of pancontig length and frequency change with the use of different alignment kernels and pseudo-energy parameters: minimap (asm10) with default parameters vs. mmseqs with null pseudo-energy. We compare the changes observed on the Hp dataset to the ones in Kp. 
In the Kp dataset the null pseudo-energies result in shorter pancontigs (lower-right panel) and a small increase in core genome fraction (lower-left panel). 
For Hp, the effect on pancontigs is quite different. More sensitive alignment with null pseudo-energies means small and intermediate pancontigs are lost, but there are slightly more long ones (top-right panel). 
It also results in a loss of low-frequency pancontigs (top-left panel) indicating that these had an homologous counterpart that with default parameters and minimap could not be merged due to high-divergence. Using a more sensitive aligner with null pseudo-energies means the core pangenome fraction increases considerably.  
      % TODO: add mention in SI
      % QUESTION: this is more or less explicit in SI? Should I expand more?
      % Liam: I think there are slightly more long pancontigs with mmseqs/null in Hp from looking at figure so seems worth saying
}

\begin{figure}[tb]
      \includegraphics[width=.7\textwidth]{figs_response/hp_vs_kp.pdf}
      \caption{frequency and length distribution (respectively left and right panels) of pancontigs for the \textit{Helicobacter pylori} and \textit{Klebsiella pneumoniae} datasets (top and bottom). For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter values and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange). Distributions are weighed by pancontig length, so that the sum of all bins heights is equal to the total pangenome graph size.}
      \label{fig:kp-vs-hp}
\end{figure}

\reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be ``left out'' in between pancontigs?}
\response{No, every part of the sequence of each genome must be included in some pancontig. 
It might not necessarily appear in the pancontig consensus sequence if it is present in the alignment as a short ($< \Lthr$) insertion, but it will nonetheless be present in the alignment. 
In the particular case in which merging two pancontigs would leave overhangs shorter than $\Lthr$, these are included in the alignment as flanking insertions. 
This is now explained in SI section {\SIgraph}, and mentioned in the main text in section IB (line 125).
In fact, we used the fact that genomes can be reconstructed exactly from the pangraph to test and debug the tool. }

@@

\reviewer{R2.9}{About data accessibility
      \begin{enumerate}
            \item i don't see how to get the list of accessions for section IIB? Saying they were used in the Ding et al paper is not quite enough for me to be sure which ones you used.
            \item for the graph marginalisation section, could the accessions of the strains used be in a supplementary file as well as on github? Github repos are great, but they can be deleted.
      \end{enumerate}}
% \Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs. I could create some excel spreadsheet with accession numbers that we could add to the submission?}\\
% \Liam{for sure that's all they want (most biologists don't like json!)}
\response{We now included in the submission two csv files. \texttt{accession\_numbers.csv} contains the list of accession numbers of all isolates in our dataset for all species considered. \texttt{marginalize\_pairs.csv} contains instead a list of all the pairs used in the graph marginalization section.
      % TODO: archive repository state.
}

\subsubsection*{Minor Issues}

\reviewer{R2.1}{line 13 - it would be good to give a reference for the statement that even closely related genomes can differ widely in gene content.
I don't know what the first paper to show this was, the earliest i am aware of is from Eduardo Rocha's team, which annpoyingly i can't locate right now, but Figure 2 here shows it \url{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1008866} \cite{touchon2020phylogenetic} - you can be very close in patristic distance and have very different gene repertoire.
Feel free to replace with a better reference, this si just meant to be helpful.}
\response{We thank the reviewer for the suggestion. We now cite the suggested article, along with \cite{touchon2009organised,doolittle2009origin}.}

\reviewer{R2.2}{Why on earth is the ``get an MSA of a block'' option in pangraph called ``polish''?}
\response{In our method pancontig merges are performed using only the consensus sequence as a guide. This makes the merging procedure very fast, but can introduce minor inconsistencies and artifacts with insertions in the alignment as insertions are not aligned to each other in the procedure.
To remove these inconsistencies, after building the graph the user can optionally decide to ``polish'' pancontig alignments with the \texttt{polish} command. This command takes a graph as input and produces a graph with the same topology, but with improved alignments. In practice for each block it collects all of the sequences and uses MAFFT \cite{katoh2002mafft} to perform a multiple sequence alignment, obtaining an improved consensus and more consistent insertion pattern.\\
This command was not mentioned in the paper. This has now been amended, and we briefly described in in section IB in the main text and in SI sect.~{\SIalgo}B.}

\reviewer{R2.3}{line 48: ``Pancontigs are connected by an edge if they are syntenic on at least one input sequence;'' - i think this is a bit of an abuse of modern language. The word syntenic is generally used to mean two genes/whatever have order conserved on multiple chromosomes (although some googling tells me that in classical genetics it meant things were on the same chromosome). Anyway, my understanding is pancontigs are connected by an edge if one of the input genomes is spelled out by going from one to the other. So it's about adjacency rather than synteny.}
\response{We agree with the reviewer's observation. We have changed ``syntenic on'' to ``adjacent in''.}

\reviewer{R2.4}{Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000}
\response{The value $N=100$ is the standard value used in the rest of the simulations (e.g. fig.~3), but as the reviewer pointed out this value is varied for the results showed in fig.~2 and the statement was confusing. We modified it removing the value of $N$ and referring to SI table~I for the standard value of simulation parameters.}

@@

\subsection*{Reviewer 3}

\reviewer{R3}{The article ``PanGraph: scalable bacterial pan-genome graph construction'' reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure. Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose. The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.\\
      For instance the authors state: ``Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.''. There are many pan-genome analysis tools which treat the pan-genome as a graph and generate a PGG. There are two primary differences in these existing tools: those which generate a graph at the nucleotide level with nodes representing basepairs (sometimes nonbranching runs of nodes are collapsed) and those where nodes represent larger features such as genes.\\
      PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.}

\response{
      We apologize if our previous discussion of the literature was incomplete. It was not our intention to downplay prior work.
      We thank the reviewer for highlighting additional important work and aspects that we have now added to better contextualise our work.
      We have rewritten the introduction to reflect previous work in this field and added to the discussion to further highlight previous and alternative approaches. We hope that these edits address the concerns of the reviewer.
      % Liam: I think good to say in more detail what has been added here. 
      Specifically, we have
\begin{itemize}
    \item The sentence in the abstract mentioned by the reviewer has been replaced by a general summary of the need to and challenges in understanding evolution of genome structure.
    \item We extended the discussion of other methods in the introduction.
    In particular we extended the paragraph on methods based on de Bruijn graphs, mentioning also compacted generalization and the fact that tools exist that are able to build graphs containing thousands of isolates. We now cite more tools and put more emphasis on their applications, restricting it to tools specific for microbial genomics (e.g. we did not include PGGB). We removed the previous claim that homology detection was limited by the k-mer size, and cite tools that can be used to extend homology detection through small sequence modifications.
    \item We expanded in the same way the paragraph on homology-based and gene-based approaches, in which we now discuss more tools putting emphasis on their applications.
    \item We added a paragraph in the discussion in which we tried to emphasize how \textit{PanGraph}, as the reviewer remarks, treats nucleotide-level variation and large-scale structural variation in different ways, thereby facilitating separate analyses of these aspects. 
\end{itemize}
}
% \Marco{Tentative extension - to be updated depending on how we modify the intro/discussion:\\

% To amend this we performed the following changes:
% \begin{itemize}
%     \item The sentence in the abstract mentioned by the reviewer has been removed and modified.

%     \item We extended the discussion of other methods in the introduction.\\
%     In particular we extended the paragraph on methods based on de Bruijn graphs, mentioning also compacted generalization and the fact that tools exist that are able to build graphs containing thousands of isolates. We now cite more tools and put more emphasis on their applications, restricting it to tools specific for microbial genomics (e.g. we did not include PGGB). We removed the previous claim that homology detection was limited by the k-mer size, and cite tools that can be used to extend homology detection through small sequence modifications.\\
%     We expanded in the same way the paragraph on homology-based and gene-based approaches, in which we now discuss more tools putting emphasis on their applications.

%     \item We added a paragraph in the discussion in which we tried to emphasize how \textit{PanGraph} compares to gene-based and k-mer based PGG methods.\\
%     In essence, methods based on de Bruijn graph have the advantage of being fast since they do not require alignments. They also do not necessarily require fully assembled genomes. However in these models the graph encodes in the same way nucleotide-level variation and large-scale structural variation, and specific post-processing of the graph is needed to distinguish the two. Conversely, PanGraph requires fully-assembled genomes and is potentially slower, but its pangenome graph representation provides a natural separation between small sequence differences (mutations or small indels) and big structural changes (gene gain/loss, inversions, rearrangements).\\
%     Gene-based pangenome graphs have instead the advantage of being smaller in size (a node corresponds to a gene) and they can include more divergent sequences, since homology detection can be performed on protein sequences. Differently from homology-based methods like PanGraph however, they are often sensitive to annotation errors.
% \end{itemize}
% }

\reviewer{R3}{An obvious lack in PanGraph is an attempt to map existing annotations back onto the generated PGG which is a necessity for usefulness to biologists. The mechanism for this is obvious but providing a useful user interface might be tedious.}
\response{We thank the reviewer for this important suggestion and appreciate the need to include a way to map annotations back onto the pangenome graph. 
We created a helper script that allows a mapping of pancontig information onto a set of annotations in a \texttt{gff} file. 
In default usage the pancontig location(s) of each feature gets added as a new `attribute', allowing a quick connection between the generated pangenome graph and existing annotations. 
The script and instructions on its usage can be found at \url{https://github.com/liampshaw/pangraph\_annotations}. 
In the future, we hope to extend this further to enable more direct biological interpretation of what the pancontigs code for. 
Integrating a fully featured annotation into pangraph was not feasible during the 30 day revision window, but is on our list of future milestones. 
}

\reviewer{R3}{
      This statement: ``While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging. Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017).'' appears to be incomplete at best. De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.\\
      One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
      \begin{itemize}
            \item \cite{angiuoli2011improving}
            \item \cite{angiuoli2011mugsy}
      \end{itemize}
      While the underlying approach is quite different, a very similar PGG to PanGraph is produced by a different method with similar nodes represented by multiple sequence alignments and edges with the same meaning:
      \begin{itemize}
            \item \cite{sutton2021pan}
            \item \cite{chan2015novel}
      \end{itemize}}
\response{
We apologize and we modified the sentence in question. We also extended the discussion of de Bruijn graphs methods in the introduction (see answer above).\\
We thank the reviewer for pointing out these important contributions that were not previously discussed in the paper.
}

\reviewer{R3}{I believe this statement: ``It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.'' is the essential step in the process and perhaps deserves more focus. It captures two important issues: aligning orthologs not paralogs and keeping the fracturing of the PGG nodes to a minimum. The pseudo-energy captures the alignment length (longer is better for orthologs versus paralogs) but does not appear to capture the surrounding context of the iterative PGGs. For initial alignments of single genomes to single genomes this doesn't matter for the most part because there is a single pancontig for each each genome and PGG context is mostly captured in the alignment. I am not so sure this is true as the PGG nodes fracture and become much smaller. When choosing between competing mergers of paralogs/orthologs the context of the PGG and other possible mergers is important if the pancontigs have been reduced so that they contain only repetitive sequence without unique flaning sequence. I think this is a very important point as distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it. This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.}
% \Liam{I am not sure I agree that constructing a pangenome graph means one must distinguish orthologs from paralogs. The structure is still useful even if this isn't done. Maybe there is a way of doing something of what they propose with simulations, but I suspect this isn't necessary for the editor to accept the paper\ldots}\\
% \Marco{I agree, I think it's a different design choice. We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph. One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}\\
\response{
This is an important point that needed more discussion in the paper than we originally provided. 
\textit{PanGraph} in its current state does not explicitly distinguish between orthologs and paralogs when constructing the graph. 
This is a deliberate design choice: PanGraph is intended to facilitate the investigation of genome evolution and two very similar paralogs represent a recent duplication or acquisition that we want to capture in the graph structure. 
Whether paralogs are merged depends on their similarity and the parameters of pangraph allow tuning this thresholds.

But we agree from experience that these can sometimes complicate the interpretion of the pangenome graph, in particular in presence of high copy-number mobile elements.  
Currently \textit{PanGraph} supports two different ways of controlling paralog splits.
The first one is based on sequence divergence as mentioned above.
By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
The second method relies instead on information on genomic context. The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag. 
When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context. 
These duplicated pancontigs are then split and absorbed in their context.
In addition, pangraph can export a graph omiting all duplicated blocks, which is a simplified version of the graph often useful for visualization.

@@

In the future we plan on providing a dedicated command to perform a more robust form of paralog splitting, which combines both sequence divergence information and context information. This command would build a phylogenetic tree from the alignment of each duplicated pancontig to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context. This functionality is not however not yet implemented in the current version of \textit{PanGraph} and we see it as substantial extra work.\\

We appreciate the reviewer's perspective and have added SI sect.~{\SIalgo}D to include discussion of these considerations.

% TODO: confirm with Nick the paralog split algorithm.
}

### 2023-03-10 reviewer_response.tex

\reviewer{R3}{In the pseudo-energy discussion a further parameter is mentioned: ``In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.''. The value chosen for this parameter should be explicitly stated along with a rationale for the choice. It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.}
\response{The minimal size parameter, now named $\Lthr$, is set to 100 bp in PanGraph. This parameter controls the minimal size of pancontigs. In general, this parameter sets a scale separation between structural variation that is encoded in the graph structure or in the pancontig alignment. 
For instance, an insertion could appear either as a separate pancontig, or as an insertion in the pancontig alignment, depending on its size. 
In practice this parameter is used in three ways:
      \begin{itemize}
            \item In addition to the pseudo-energy constraint, merges are refused if they have alignment length shorter than $\Lthr$.
            \item During merging, if the alignment contains long ($> \Lthr$) indels, these will cause a split in the merged pancontigs and will be treated as separate pancontigs.
            \item If a merger of two pancontigs would produce flanking overhangs shorter than $\Lthr$, these are not saved as separate pancontigs but are included in the alignment as flanking indels.
      \end{itemize}
      We choose a value of 100 bp in our analysis (and as standard value) to be able to detect variations at the gene scale, while encoding variations of up to few tens of bps in the alignments. This value can however be tuned by the user depending on the specific needs.\\
      We appreciate the need to better explain the role of this parameter in the merging procedure and in the structure of the resulting graph. To this end we extended its definition in section IB in the main text, and give a more extended discussion in SI section~\SIalgo B and \SIalgo C.\\
      % TODO: Homogenization in the main text with discussion on precision. Extend.
}

\reviewer{R3}{How are pancontig consensus sequences determined and updated? I could not find any details about this. This is not a trivial operation. My reading of the article seems to imply that a complete multiple sequence alignment is not done for each merger. Does the consensus sequence degenerate iteratively with mergers? For the simulated data how accurate are the final pancontig consensus sequences?}

\response{This point was raised by other reviewers, and we appreciate the need to make this part of the algorithm clearer. 
To this end we added section {\SIalgo}B in the SI, with a detailed explanation of how the merging procedure works.\\
In essence, during graph merging alignment is only performed between the consensus sequences of the merged pancontigs. 
We avoid performing a multiple sequence alignment during merging to keep the algorithm scalable. 
However we do update the consensus of the pancontig at each merge using the majority rule on the merged alignment.\\
The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an excessive amount of resources. 
This separates efficiently the problem of resolving the large-scale and nucleotide-level diversity between sequences.\\
However after the graph has been built, alignments can be refined with \textit{PanGraph}'s \texttt{polish} command. 
This command performs a multiple sequence alignment on all the sequences in each pancontig and updates the consensus sequence accordingly. In our experience, the alignments appear reliable throughout this process, although we have not explicitly quantified it from our simulations. However, for cases where the alignments are of interest we recommend running `polish' to perform MSA for pancontig consensus sequences (as in the documentation) to ensure that the final alignments are as accurate as possible.  
}

@@

\reviewer{R3}{It is not clear to me what Table 1 shows. Is this published data from PanX or data from PanGraph? Either way the value of 0.7 Mbp for the E. coli core genome length seems implausibly low for high quality complete genomes. The authors should justify this. The authors should also include analysis on the calculated core genome length versus actual core genome length for the simulated data.}
\response{
We thank the reviewer for pointing out this issue with the E.~coli dataset.\\

Concerning the origin the data, our datasets are taken from the PanX pipeline (\url{https://pangenome.org/}). They were not included in the original publication but are part of a later update, dated March 2018. Values displayed in table I are extracted from the PanX analysis. This is now made more explicit in the text.\\

Concerning the core genome size, the unusually small value of 0.7 Mbp is due to the fact that we used a strict definition of core genes as single-copy genes that are present in every isolate. This definition is very sensitive to annotation errors.\footnote{
      For example in \cite{sutton2021pan} using a similar strict definition the authors report finding $\sim$700 core orthologous gene clusters for an E.~coli dataset of $\sim$1000 genomes (see table 2 in their publication), compatible with our estimate of 0.7 Mbp. They impute this to misannotations, since after refining the annotations using structural information from the pangenome this number increases to $\sim$1500 genes.
} If we measure the size of the soft-core genome with a frequency threshold of 95\% we find a more reasonable size of 2 Mbp. We now report both of these values in table~I.\\
A careful investigation of the dataset also revealed that a pair of records have been subsequently removed from RefSeq, namely \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_020518.1?report=genbank}{NC\_020518} (an artificial genome) and  \href{https://www.ncbi.nlm.nih.gov/nuccore/NC_017663.1?report=genbank}{NC\_017663}. However these are not at the origin of the small size of the core genome. In fact by removing these isolates the number of core genes only shows a very limited increase from 778 to 822, and the number of soft-core genes decreases from 2081 to 2069.\\

To make this more clear we updated part of the text in section IIB, the caption of table~I and SI sect.~{\SIdataBenchmark}A.
}

\reviewer{R3}{The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.}
% \Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
\response{
      We appreciate the importance of the point raised by the reviewer.
      We agree with the reviewer that this post-processing smoothing of the graph produced by of \textit{PanGraph} would be a valuable addition, and we hope to include such further post-processing refinements in subsequent versions. 
      This sort of refinement would allow alignments of adjacent edges of pancontigs to be harmonized, potentially also removing very small pancontigs when these could be appropriately absorbed in the alignment of larger ones. We hope to continue to expand the downstream post-processing of \textit{PanGraph} data structures in future work. 
      For small scale variation within pancontigs, we already have the {\it polish} command. 
}

\reviewer{R3}{Overall the paper is well written. I like the basic methodology and the attention to using robust efficient existing algorithms and combining them to keep efficiency in mind. I am very concerned about the potential fracturing of the PGG where unfortunately the devil is in the details and am not convinced that the fracturing will not prevent some limit on the number of genomes which can be used for a pangenome.}
% \Marco{if the fracturing derives from a real structural property of pangenomes (i.e. ISs that really move around and split long blocks) this might be an intrinsic limit that derives from the pangenome, I wouldn't consider it a problem with PanGraph.}\\
\response{
      We thank the reviewer for their appreciation of the paper and of our methodology.\\
      We agree with the reviewer that fragmentation of the graph is an important concern, and that it can constitute a limitation in the number and diversity of genomes that can meaningfully be added in the pangenome graph. 
      Within the size limit of our datasets, in the paper we investigated fragmentation in sect.~2B (``\textit{scaling with dataset size}''). 
      In particular in fig.~5 we quantify how the size and number of pancontigs vary as a function of the number of isolates included in the graph. 
      Here we observed good scaling properties of the pancontig size as more isolates are added (much slower than $1/n$).
      This is true in particular for core pancontigs. 
      However, if the size of contigs decreases to a degree that it is comparable to the minimal length threshold $\Lthr$, we expect the graph to be no longer very useful.

      %Irrespective of its origin, a possible way to detect excessive fragmentation and check for consistency in the graph structure is to visualize the distribution of pancontig size in the graph (see for instance fig.~\ref{fig:kp-vs-hp} lower right panel). Excessive fragmentation in the graph would be detectable as most of the sequence being condensed in pancontigs having size close to $\Lthr$.\\
      % TODO: show distributions for pangenome graphs with 500/1k isolates? Mention times?
      % TODO: add this discussion in the text?
      % TODO: mention the partition entropy? This could be an argument against artificial fragmentation.
      % TODO: maybe add in the SI the block frequency distribution for different species/kernels?
}
